---
title: "Logistic Regression: Horse Racing"
author: "Kyubin Im, Matthew Murnane, and Ethan Newcomb"
format: revealjs
---

```{r include=FALSE}
library(tidyverse)
library(ggplot2)
library(gridExtra)
library(readr)
library(ggthemes)
library(forcats)
library(knitr)
library(kableExtra)
```

```{r}

runs <- read_csv("..\\data\\runs.csv")
races <- read_csv("..\\data\\races.csv")

# Join --------------------------------------------------------------------

whole <- inner_join(runs, races, by="race_id")

# Changing Data Classes ---------------------------------------------------

whole$race_id <- as.character(whole$race_id)
whole$horse_no <- as.character(whole$horse_no)
whole$horse_id <- as.character(whole$horse_id)
whole$won <- as.logical(whole$won)
whole$horse_ratings <- as.character(whole$horse_ratings)
whole$draw <- as.character(whole$draw)
whole$jockey_id <- as.character(whole$jockey_id)
whole$trainer_id <- as.character(whole$trainer_id)
whole$race_no <- as.character(whole$race_no)
whole$surface <- as.character(whole$surface)
whole$race_class <- as.character(whole$race_class)
whole$time7 <- as.numeric(whole$time7)
whole$sec_time7 <- as.numeric(whole$sec_time7)
whole$distance <- as.character(whole$distance)
whole$result <- as.character(whole$result)

position_cols <- paste0("position_sec", 1:6)
place_cols <- paste0("place_combination", 1:4)
wincomb_cols <- paste0("win_combination",1:2)

for (col in c(place_cols, position_cols)) {
  whole[[col]] <- as.character(whole[[col]])
}
```

## The Data

- A horse per race is the observation
- 79,447 rows and 73 columns

```{r}
kable(head(whole)) %>%
  kable_styling(font_size = 15, full_width = FALSE)
```

## Some Variables

- `horse_country`: country of origin of this horse
- `declared_weight`: declared weight of the horse and jockey, in lbs
- `draw`: post position number of the horse in this race
- `position_sec`[i]:  position of this horse (ranking) in $i^{th}$ section of the race
- `time`[i]: time taken by the horse to pass through the $i^{th}$ section of the race (sec)
- `win_odds`: win odds for this horse at start of race
- `jockey_id`: unique identifier of the jockey riding the horse in this race

## Levels of Categorical Variables

```{r}
level_counts <- data.frame(levels=sapply(sapply(whole[,sapply(whole, is.character)], unique), length))

level_counts %>% 
  arrange(desc(levels)) %>% 
  kable(
    format = "html",
    col.names = c("Variable", "Number of Levels")) %>%
  kable_styling(full_width = FALSE) %>%
  scroll_box(height = "600px") 
```

